---
title: "Assignment 6: ATAC-seq"
output:
  github_document:
    toc: true
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Assignment overview

*Today we will be looking at a differential ATAC-seq dataset between cells treated with an anti BAF protac and control (untreated) cells. The cell type is HAP1, a cancer cell line with a near-haploid genome. We will use this dataset to explore differential analysis. *

*The GEO entry is located here, where you can read more about the experiments: https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE148175 *

*This is the paper: https://www.nature.com/articles/s41588-021-00777-3 *

*"Acute BAF perturbation causes immediate changes in chromatin accessibility"*



# Part 0: Getting ready 

```{r}
#install any of these you might not have already
library(ggplot2)
library(edgeR)
library(reshape)
library(GenomicRanges)
library(csaw)
library(Biostrings)
```


```{r}
#download the data
atacSeqData = read.table(textConnection(readLines(gzcon(url("ftp://ftp.ncbi.nlm.nih.gov/geo/series/GSE148nnn/GSE148175/suppl/GSE148175_count_matrix_raw_atac_BRM014_ACBI1.csv.gz")))), 
                      sep=",", stringsAsFactors = FALSE, header = TRUE)
```


```{r}
#create a sample metadata data.frame
samples = data.frame(ID = names(atacSeqData)[2:ncol(atacSeqData)])
samples$replicate = gsub("(R[12])_([0-9]+[minh]+)_(.*)$","\\1",samples$ID)
samples$timeName = gsub("(R[12])_([0-9]+[minh]+)_(.*)$","\\2",samples$ID)
samples$treatment = gsub("(R[12])_([0-9]+[minh]+)_(.*)$","\\3",samples$ID)
samples$treatment[samples$treatment=="N"]="BRM014"
samples$time= as.numeric(gsub("[a-z]*","",samples$timeName))
samples$time[grepl("min",samples$timeName)]=samples$time[grepl("min",samples$timeName)]/60
```



# Part 1: understanding the experiment

*Now using `samples` make a plot showing the experimental design, with time on the x axis, treatment on the y axis, and one plot on the left and one on the right for the two replicates (e.g. using `facet_grid`).*

### `#?#` *Make the above plot. Each point should represent one of the samples.  - 1 pt*
```{r}
#here, if the point is there, it means such a sample exists, if absent it means that there is no such sample
ggplot(data = (samples), mapping = aes(x = timeName, y = treatment)) + 
  geom_point() +
  facet_wrap( ~replicate)
```

*In this study, one of the things they were comparing was BRM014 to DMSO. The drug BRM014 is dissolved in DMSO, so DMSO alone is the appropriate control to gauge the effect of BRM014.*

### `#?#` *Can we compare BRM014 to DMSO across all time points? Why/why not?  - 1 pt*
No, we cannot compare the two across all timepoints, This is because DMSO does not have time points for 10 mins, 30 mins and 6 hours, where as BRM014 has values taken for all time points.

# Part 2: QC

*With most genomics data, it is important both that samples have sufficient coverage, and that the samples have similar coverage. Either case can lead to underpowered analysis, or misleading results. Calcualte the read coverage for each sample. *

### `#?#` Make a plot with read coverage on the y-axis (total number of reads) and the samples on the x-axis. - 3 pt*

```{r}
# there are many ways you could do this; one of which is using the melt/cast functions from reshape
library(tidyverse)

atacSeqData_reads_samples <- melt(atacSeqData)
atacSeqData_reads_samples_total <- aggregate(atacSeqData_reads_samples$value, by=(list(ID = atacSeqData_reads_samples$variable)), FUN = sum)

atac_reads_joined <- left_join(atacSeqData_reads_samples_total, samples, by.x =atacSeqData_reads_samples_total$ID, by.y = samples$ID )

ggplot(data = (atac_reads_joined), mapping = aes(x = ID, y = x, colour = replicate)) + 
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))


```

### `#?#` *Which sample has the most coverage? - 0.5 pt*
```{r}
#sample with the highest coverage is R1_24h_DMSO
atac_reads_joined[which.max(atac_reads_joined$x), ]
```

### `#?#` *Which sample has the least? - 0.5 pt*
```{r}
#sample with min reads is R1_6h_control
atac_reads_joined[which.min(atac_reads_joined$x), ]
```

### `#?#` *What is the % difference between the max and min (relative to the min)? - 0.5 pt*
```{r}
percent_diff = ((max(atac_reads_joined$x) - min(atac_reads_joined$x))/min(atac_reads_joined$x))*100

percent_diff

```

*In cases where samples have vastly different coverage, you can potentially down-sample the higher-coverage samples. Sometimes, throwing out the data in this way can also introduce new problems, so we're going to stick with the data we have.*

*For this assignment, we will look only at BI_protac vs control data. *

### `#?#` *Create a new data.frame containing only the BI_protac and control samples - 1 pt*
```{r}
atacSeqData_with_treat <- mutate(atacSeqData_reads_samples, treatment=sapply(str_split(as.character(atacSeqData_reads_samples$variable), "_", n=3), function(x) x[[3]][1]) )


atac_reads_filtered <- atacSeqData_with_treat[atacSeqData_with_treat$treatment == "BI_protac" |atacSeqData_with_treat$treatment == "control", ]

head(atac_reads_filtered)


```

### `#?#` *For this subset, calculate the counts per million reads (CPM) for each sample - 2 pt*
```{r}
atac_reads_filtered$cpm <- cpm(atac_reads_filtered$value)
head(atac_reads_filtered)
```


### `#?#` *Plot the kernel density estimate for CPM (x axis). 1 curve per sample, different colours per curve. - 1 pt*

```{r}
 ggplot(data = atac_reads_filtered, aes(x = cpm, color = variable)) +
  geom_density() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### `#?#` *Plot the kernel density estimate for log(CPM+1) (x axis), coloured as before - 1 pt*

```{r}
atac_reads_filtered <- atac_reads_filtered %>% mutate(log_pseudo = log(atac_reads_filtered$value + 1))
head(atac_reads_filtered)

ggplot(data = atac_reads_filtered, aes(x = log_pseudo, color = variable)) +
  geom_density() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

### `#?#` *Why do you think log-transforming is usually performed when looking at genomics data? What about adding 1 before log transforming? - 2 pt*

The reason we look at log transformed data in genomics is because there can be drastic differences in the counts values, from being very small to very large. These drastic differences can make it difficult to compare relationships between values and identify trends. We use the addition of 1 because if the cpm =0, you can't take the log of 0, so if you add 1, you're taking the log of 1 as opposed to 0, so you don't need to deal with undefined values.


### `#?#` *Some regions have very large CPMs. Inspect the peaks for which CPM>400. What do you notice about them? 3 pt*
```{r}
atac_reads_filtered_high_cpm <- atac_reads_filtered[atac_reads_filtered$cpm > 400, ]
atac_reads_filtered_high_cpm

table(atac_reads_filtered_high_cpm$region)

#they all appear to be from the same 4 regions 
#chrM:12425-14404 chrM:14788-15943     chrM:198-854   chrM:2059-2771   chrM:2868-3973 

```

*Normally, we would remove some of these regions before continuing (and would redo the above steps). Since this is an assignment, we will continue with the data as-is.*

*Often a good first step is to see if the data look good. One way to do this is by seeing whether or not the signals in each sample correlate with each other in ways you expect.*

### `#?#` *Calculate the pairwise correlations between log(CPM+1)s for the samples and plot them as a heatmap (samples x samples) - 3 pt*
```{r}
frame <- data.frame(atac_reads_filtered$variable, atac_reads_filtered$log_pseudo, atac_reads_filtered$region)
#frame
frames2 <- spread(frame, key = atac_reads_filtered.variable, value = atac_reads_filtered.log_pseudo)
frames2 <- frames2 %>% remove_rownames %>% column_to_rownames(var="atac_reads_filtered.region")

cor_mat <- cor(frames2)
cor_mat
heatmap(cor_mat)
melt_cor <- melt(cor_mat)
melt_cor
ggplot(melt_cor, aes(x = X1, y = X2, fill = value)) + geom_tile() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### `#?#` *What do you expect the correlations between replicates to look like? Is that what you see? - 2 pt*



*It is common to exclude some regions from analysis. For instance, we won't be able to robustly identify those that are differential but have low coverage even if they are truly differential, so there is no point testing these. We will also remove mitochondrial regions, a common contaminant of ATAC-seq data.*


### `#?#` *Filter your data, retaining only regions where the average counts per sample is greater than 10, and also remove mitochondrial regions - 3 pt*
```{r}
frame3 <- data.frame(atac_reads_filtered$variable, atac_reads_filtered$value, atac_reads_filtered$region)
#frame
frames4 <- spread(frame3, key = atac_reads_filtered.variable, value = atac_reads_filtered.value)

frames4 <- frames4 %>% remove_rownames %>% column_to_rownames(var="atac_reads_filtered.region")
frames5 <- frames4[ ,colMeans(frames4[sapply(frames4,is.numeric)]) > 10]


frames5 <-  mutate(frames5, chr = sapply(str_split(as.character(rownames(frames5) ), ":"), function(x) x[[1]][1]) )
frames5_filtered <- frames5[frames5$chr != "chrM", ]


```

### `#?#` *How many peaks did you have before? How many do you have now? - 1 pt*
```{r}
frames5_filtered <- rownames_to_column(frames5_filtered, var = "region")
frame_5_melt <- melt(frames5_filtered)
frame_5_melt$cpm <- cpm(frame_5_melt$value)
frame_5_melt$log_pseudo <- log(frame_5_melt$cpm + 1)
frame_5_melt

ggplot(data = frame_5_melt, aes(x = log_pseudo, color = variable)) +
  geom_density() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

#before, there was 1 central peak. Now there are two peaks

```


# Part 3: Differential ATAC

*We want to know what regions are differentially accessible between BI_protac and the control.* 

*Today, we're going to use edgeR, which is designed for RNA-seq, but works well on ATAC-seq as well. The user guide is here:* https://www.bioconductor.org/packages/release/bioc/vignettes/edgeR/inst/doc/edgeRUsersGuide.pdf


### `#?#` *Make a count matrix called `countMatrix` for the BI_protac and control samples, including only the peaks we retained above - 2 pt*
```{r}
#remove non-numeric values from the matric
countMatrix <- frames5_filtered %>% remove_rownames %>% column_to_rownames(var = "region")

                                                                          
countMatrix$chr <- NULL
#show the matrix
head(countMatrix)
```

*EdgeR is exceptionally versatile, with many different options for analysis. Today, you're going to use the GLM-quasi-likelihood approach to calculate differential accessibility. We are providing the first example analysis below, which you can modify in subsequent steps. You will need to understand what the steps do, so read the appropriate documentation. *
```{r, eval = FALSE}
curSamples = samples[match(names(countMatrix), samples$ID),];
y = DGEList(counts=countMatrix, group=curSamples$treatment)
y = calcNormFactors(y)
designPaired = model.matrix(~curSamples$treatment + curSamples$timeName)  
# we are using timeName here to make sure that time is treated as a categorical variable. Had we more time points it might make sense to treat time as a value.
y = estimateDisp(y, designPaired)
fitPaired = glmQLFit(y, designPaired)
qlfPairedTime6vs24 = glmQLFTest(fitPaired, coef=3) 
qlfPairedTreatControlvsProtac = glmQLFTest(fitPaired, coef=2)
allDEStatsPairedTreatControlvsProtac = as.data.frame(topTags(qlfPairedTreatControlvsProtac,n=nrow(countMatrix)))
allDEStatsPairedTreatControlvsProtac$region=row.names(allDEStatsPairedTreatControlvsProtac)
allDEStatsPairedTime6vs24 = as.data.frame(topTags(qlfPairedTime6vs24,n=nrow(countMatrix)))
allDEStatsPairedTime6vs24$region=row.names(allDEStatsPairedTime6vs24)
```
*While the differential analysis has been done in this case, before we look at the results, we are going to check if the data appear to be normalized correctly. Also include a loess line of best fit, and the line y=0.*

### `#?#` *Make an MA plot for allDEStatsPairedTreatControlvsProtac -2pt*
```{r}
ggplot(data = allDEStatsPairedTreatControlvsProtac, mapping = aes(x = logCPM, y = logFC)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0), colour = "blue", size = 1.0)  +
  stat_smooth(method = "loess", se = FALSE, colour = "red")
  


```

### `#?#` *Make an MA plot for allDEStatsPairedTime6vs24 - 1 pt*
```{r}
ggplot(data = allDEStatsPairedTime6vs24, mapping = aes(x = logCPM, y = logFC)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0), colour = "blue", size = 1.0) +
  stat_smooth(method = "loess", se = FALSE, colour = "red")
```

*Now we're going to test loess normalization instead.* 


### `#?#` *Perform the same differential peak analysis using loess regularization. - 1 pt*
```{r}
#Note: the Bioconductor package csaw implements loess regularization in a way that is compatible with edgeR
## Tip: use the csaw library to implement the loess regularization
thing <- normOffsets(y)

designPaired2 = model.matrix(~curSamples$treatment + curSamples$timeName)  
# we are using timeName here to make sure that time is treated as a categorical variable. Had we more time points it might make sense to treat time as a value.
thing = estimateDisp(thing, designPaired2)
fitPaired2 = glmQLFit(thing, designPaired2)
qlfPairedTime6vs242 = glmQLFTest(fitPaired2, coef=3) 
qlfPairedTreatControlvsProtac2 = glmQLFTest(fitPaired2, coef=2)
allDEStatsPairedTreatControlvsProtac2 = as.data.frame(topTags(qlfPairedTreatControlvsProtac2,n=nrow(countMatrix)))
allDEStatsPairedTreatControlvsProtac2$region=row.names(allDEStatsPairedTreatControlvsProtac2)
allDEStatsPairedTime6vs242 = as.data.frame(topTags(qlfPairedTime6vs242,n=nrow(countMatrix)))
allDEStatsPairedTime6vs242$region=row.names(allDEStatsPairedTime6vs242)

```

### `#?#` *Make the same two MA plots as before, but this time using the loess normalized analysis - 1 pt*
```{r}
ggplot(data = allDEStatsPairedTreatControlvsProtac2, mapping = aes(x = logCPM, y = logFC)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0), colour = "blue", size = 1.0)  +
  stat_smooth(method = "loess", se = FALSE, colour = "red")

ggplot(data = allDEStatsPairedTime6vs242, mapping = aes(x = logCPM, y = logFC)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0), colour = "blue", size = 1.0)  +
  stat_smooth(method = "loess", se = FALSE, colour = "red")
```

### `#?#` *What was the first normalization method? What changed in the MA plots? Which analysis do you think is more reliable and why? - 4 pt*

original version was from calcNormFactors(). 
The thing that changed was the loess line

# Part 4: GC bias

*Next, we will look at potential GC bias in the data. We will again use bioconductor *

### `#?#` *Convert the region IDs to a GRanges object - 3 pt*
```{r}
#note that the names of your peaks are of the format <chr>:<startPos>-<endPos>
## Tip: lookinto the GenomicRanges documentation 
#print(frames5_filtered$region)

rownames(frames5_filtered) <- frames5_filtered$region

region_info.df <- data.frame(do.call("rbind", strsplit(as.character(frames5_filtered$region), "[:-]")), row.names = frames5_filtered$region )
colnames(region_info.df) <- c("chr", "startPos", "endPos")

region_info.df

ir <- IRanges(start=as.numeric(region_info.df$startPos), end=as.numeric(region_info.df$endPos))
gr <- GRanges(seqnames = region_info.df$chr, ranges = ir, names = rownames(region_info.df))
gr


```


### `#?#` *Extract the genomic DNA sequences for each peak using hg38 - 3 pt*
*See for relevant documentation: https://bioconductor.org/packages/release/workflows/vignettes/sequencing/inst/doc/sequencing.html *
```{r}
## Tip: Use the Biostring library 
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")
#BiocManager::install("BSgenome.Hsapiens.UCSC.hg38")

library(BSgenome.Hsapiens.UCSC.hg38)
sequences <- getSeq(BSgenome.Hsapiens.UCSC.hg38, gr)

freqs <- alphabetFrequency(sequences)
gc_content <- (freqs[,'C'] + freqs[,'G'])/rowSums(freqs)
frames5_filtered$gc <- gc_content
```


*Now we will see if there's any relationship between peak CPM and GC content for each of the samples.*

### `#?#` *Create scatter plots (one per sample, e.g. using facet_wrap), including lines of best fit (GAM), where each plot shows GC content (x axis) vs CPM (y axis) for each peak (points) -2pt*
```{r}
#please limit the y axis to between 0 and 50

frame5_filtered_long <- pivot_longer( frames5_filtered, cols = c(R1_24h_control, R1_24h_BI_protac, R1_6h_BI_protac, R1_6h_control, R2_24h_control, R2_24h_BI_protac, R2_6h_BI_protac, R2_6h_control),  names_to = "sample")

ggplot(data = frame5_filtered_long, mapping = aes(x=gc, y =  value)) + 
  geom_point() + 
  geom_smooth(method = "gam", se = FALSE) +
  scale_y_continuous(limits = c(0, 50)) +
  facet_wrap(~sample)

```

### `#?#` *Repeat the above, but this time showing only the lines of best fit and all on the same plot - 2 pt*
```{r}
ggplot(data = frame5_filtered_long, mapping = aes(x=gc, y =  value, colour = sample)) + 
  geom_smooth(method = "gam", se = FALSE)
```


### `#?#` *Given this result, predict whether we will see a significant relationship between GC content and logFC in our differential peak analysis (loess-normalized). Justify your prediction. Predicting "wrong" will not be penalized, as long as your justification is correct. Don't retroactively change your answer. - 2 pt*

I predict that I will see a significant relationship between GC content and log fold change. This is because according to the above figure, we can see that from the figure higher GC content (0.6 - 0.8) appears to result in higher read couunts.



### `#?#` *Plot the relationship between GC and logFC for the loess-normalized ControlvsProtac analysis. Also include a line of best fit (blue) and y=0 (red) - 2 pt*
```{r}
gc_and_region <- data.frame(frames5_filtered$region, frames5_filtered$gc)
colnames(gc_and_region) <- c("region", "gc")

allDEStatsPairedTreatControlvsProtac2_gc <- left_join(x = allDEStatsPairedTreatControlvsProtac2, y = gc_and_region, by.y = allDEStatsPairedTreatControlvsProtac2$region, by.x = gc_and_region$region )

ggplot(data = allDEStatsPairedTreatControlvsProtac2_gc, mapping = aes(x = gc, y = logFC)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0), colour = "red", size = 1.0)  +
  geom_smooth(method = "loess", se = FALSE, colour = "blue")


```


### `#?#` *Now plot the same thing for the NON loess-normalized ControlvsProtac analysis. - 1 pt*
```{r}
allDEStatsPairedTreatControlvsProtac_gc <- left_join(x = allDEStatsPairedTreatControlvsProtac, y = gc_and_region, by.y = allDEStatsPairedTreatControlvsProtac$region, by.x = gc_and_region$region )

ggplot(data = allDEStatsPairedTreatControlvsProtac_gc, mapping = aes(x = gc, y = logFC)) + 
  geom_point() +
  geom_hline(aes(yintercept = 0), colour = "red", size = 1.0)  +
  geom_smooth(method = "loess", se = FALSE, colour = "blue")

```


### `#?#` *Was your prediction correct? Do you think we should also account for GC normalization in our differential ATAC analysis? Why/why not? - 3 pt*

Yes, my prediction was correct. I think that we should account for GC normalization in differential ATAC analysis. This is because by not accounting for it, it can have a drastic impact on the resulting fold change values.

*We will leave GC normalization as an optional exercise, and will not actually do it here.*

# Part 5: Differential analysis results

### `#?#` *Suppose we perform the analyses above, redoing the differential analysis once more with GC normalization, and also considering that we tested loess and the default normalization methods. Did we P-hack? Why or why not? - 2 pt*

P-hacking revolves around the idea that 

*Going forward, we will only use the initial analysis (**not loess normalized**)*

### `#?#` *Now considering the two comparisons (6 vs 24 hours, and protac vs control). EdgeR performed a correction for MHT, but if we want to analyze the results from both comparisons, do we need to re-adjust to account for the fact that we tested two different hypothesis sets (time and treatment)? Why/not? - 2 pt*




### `#?#` *How many differential peaks did you find (FDR<0.01). - 1 pt*
```{r}
```

### `#?#` *Make a volcano plot of the allDEStatsPairedTreatControlvsProtac, with -log10(p-value) on the y axis and logFC on the x. Colour points that are significant at an FDR<0.01. - 2 pt*
```{r}
```




### `#?#` *Plot the logCPM (x axis) by -log10(Pvalue) (y axis), again colouring by FDR<0.01. - 2 pt*
```{r}
```

### `#?#` *Do you think our initial filtering on peaks with at least 10 reads on average per sample was a good choice? Why or why not?*



*At this point there are many other follow ups you can and would do for a real differential analysis, but we leave these as optional exercises. For example:*
1. Confirming that the differential peaks look correct (e.g. CPM heatmap)
2. Confirming that peaks look differential on the genome browser
3. Looking for motif enrichment
4. Performing a GREAT analysis, including functional enrichment and assigning peaks to genes

*Knit your assignment as a github_document and submit the resulting .md and this .Rmd to your github, and complete the assignment submission on Canvas. Make sure to include the graphs with your submission. *
 



# Authors and contributions

Following completion of your assignment, please fill out this section with the authors and their contributions to the assignment. If you worked alone, only the author (e.g. your name and student ID) should be included.

Authors: Neera Patadia (79557773)

